---
title: "Stats 415: Final Project Code"
author: Andrew Mashhadi
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load Libraries

```{r, warning=FALSE, message=FALSE}

library(MASS)
library(astsa)

```

# Loading Data

```{r}

# import data 
df <- read.csv('dataset/stackexchangecounts_ts.csv')
df <- df[-c(1, 2, nrow(df)), ] # remove incomplete 2008

# first few rows of data
head(df)

# summary of data
summary(df)

# set topic to time-series
count.dat <- df$tagcount

# create time series objects for train/test split
xt <- ts(count.dat, start = c(2009, 1), frequency = 12)

trn <- count.dat[1:floor(0.9*length(count.dat))]
tst <- count.dat[(floor(0.9*length(count.dat))+1):length(count.dat)]
xt_trn <- ts(trn, start = c(2009, 1), frequency = 12)
xt_tst <- ts(tst, start = c(2021, 12), frequency = 12)

# create time plot of training data
ts.plot(xt_trn, 
        type = "l", 
        lwd = 2,
        main = "Monthly \"Time-Series\" Tagged Question Counts",
        xlab = "Year", 
        ylab = "Counts")
grid(lty="solid")
lines(xt_trn)


# acf and pacf, before trend removed
acf2(xt_trn, max.lag = 20, lwd = 2, 
     main="Series: \"Time-Series\" Tagged Question Counts")


```


# Removing trend

```{r}

# fit linear and poly degree 3
xt_trn_df <- data.frame(time=as.vector(time(xt_trn)), count=as.vector(xt_trn))
lm.fit <- lm(count ~ time, data=xt_trn_df) # linear
lm.fit.d3 <- lm(count ~ poly(time, 3), data=xt_trn_df) # poly degree 3

# summarys
summary(lm.fit)
summary(lm.fit.d3)

# compare RSS
anova(lm.fit, lm.fit.d3)

ts.plot(xt_trn, 
        type = "l", 
        main = "Monthly \"Time-Series\" Tagged Question Counts",
        xlab = "Year", 
        ylab = "Counts",
        lwd=2)
grid(lty="solid")
lines(xt_trn)
lines(list(x=time(xt_trn), y=lm.fit.d3$fitted.values), 
      lwd=2, 
      col="blue")
legend(2008.7, 220, 
       legend=c("Raw Counts", "Fitted (Deg. 3)"),
       col=c("black", "blue"), 
       lwd=2, 
       cex=0.8)

# remove trend relationship, show plot
detrended_xt_trn <- xt_trn - lm.fit.d3$fitted.values 
ts.plot(detrended_xt_trn, 
        main = "Monthly \"Time-Series\" Question Counts Detrended",
        xlab = "Year", 
        ylab = "Detrended Residuals", 
        type = "l",
        lwd=2) 
grid(lty="solid")
lines(detrended_xt_trn)



```


# Spectral Analysis

```{r, fig.height=6}

par(mfrow=c(2, 1))

# investigate trend/cycles
dat.spec <- mvspec(detrended_xt_trn, log="no", taper=0, 
                   main="Series: Detrended Training Data | Raw Periodogram | taper = 0", 
                   lwd=2)

# non-parametric
dat.spec.sm.np <- mvspec(detrended_xt_trn, 
                         kernel("modified.daniell", c(2, 2)), # not a large set of data points
                         taper=0.2,
                         log="no", 
                         main="Series: Detrended Training Data | Smoothed Periodogram | taper = 0.2", 
                         lwd=2)

# find min ar order that minimizes the training RMSE 
n <- length(detrended_xt_trn)
RMSE <- rep(0, 30)
for (k in 1:30){
  
  resid <- ar(detrended_xt_trn, order=k, aic=FALSE)$resid
  RMSE[k] <- sqrt(mean(resid^2, na.rm=TRUE))

}

plot(RMSE, type="o", xlab="p", ylab="Training RMSE", 
     main="Tuning AR(p) Fit with RMSE",
     lwd=2)
grid(lty="solid")
lines(RMSE, lwd=2)

# parametric
dat.spec.sm.p <- spec.ar(detrended_xt_trn, order=15, log="no", 
                         main="Series: Detrended Training Data | AR(15) Spectrum")
axis(1, at = seq(0, 6, by=0.5), tck = 1, col = "lightgrey", labels=FALSE, lty = "solid")
axis(2, tck = 1, col = "lightgrey", labels = FALSE, lty = "solid", tick=T)
axis(1, at = seq(0, 6, by=0.5), tick = TRUE, labels=FALSE)
axis(2, tick = TRUE, labels=FALSE)
lines(dat.spec.sm.p$freq, dat.spec.sm.p$spec, lwd=2)

# two-sided confidence intervals
cat("\nTWO-SIDED CONFIDENCE INTERVALS\n")
df <- dat.spec.sm.np$df 
U <- qchisq(.025, df) 
L <- qchisq(.975, df) 
cat("For peak at freq =", dat.spec.sm.np$freq[4], 
    "has the approximate two-sided CI: [", df*dat.spec.sm.np$spec[4]/L, ",", 
    df*dat.spec.sm.np$spec[4]/U, "]\n")
cat("For peak at freq =", dat.spec.sm.np$freq[13], 
    "has the approximate two-sided CI: [", df*dat.spec.sm.np$spec[13]/L, ",", 
    df*dat.spec.sm.np$spec[13]/U, "]\n")
cat("For peak at freq =", dat.spec.sm.np$freq[39], 
    "has the approximate two-sided CI: [", df*dat.spec.sm.np$spec[39]/L, ",", 
    df*dat.spec.sm.np$spec[39]/U, "]\n")


# one-sided confidence intervals
cat("\nONE-SIDED CONFIDENCE INTERVALS\n")
L <- qchisq(.95, df) 
cat("For peak at freq =", dat.spec.sm.np$freq[4], 
    "has the approximate one-sided CI: [", df*dat.spec.sm.np$spec[4]/L, 
    ", inf ]\n")
cat("For peak at freq =", dat.spec.sm.np$freq[13], 
    "has the approximate one-sided CI: [", df*dat.spec.sm.np$spec[13]/L,
    ", inf ]\n")
cat("For peak at freq =", dat.spec.sm.np$freq[39], 
    "has the approximate one-sided CI: [", df*dat.spec.sm.np$spec[39]/L, 
    ", inf ]\n")


```


We now visualize the annual cycle with a few example years.


```{r, fig.align='center', fig.height=9, fig.width=6}

# set up 4x1 block of plots 
months <- c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D")
par(mfrow=c(4, 1))


# setup plot 
years.f2009 <- 5
t1 <- 12*years.f2009 + 1
t2 <- t1 + 11

plot(time(detrended_xt_trn)[t1:t2], detrended_xt_trn[t1:t2], 
     type = "c", xlab = "", ylab = "", xaxt = 'n', lwd=2,
     main = "Detrended Residuals Throughout 2014") 
points(time(detrended_xt_trn)[t1:t2], detrended_xt_trn[t1:t2], pch = months, col = 1:4)

# setup plot 
years.f2009 <- 6
t1 <- 12*years.f2009 + 1
t2 <- t1 + 11

plot(time(detrended_xt_trn)[t1:t2], detrended_xt_trn[t1:t2], 
     type = "c", xlab = "", ylab = "", xaxt = 'n', lwd=2,
     main = "Detrended Residuals Throughout 2015") 
points(time(detrended_xt_trn)[t1:t2], detrended_xt_trn[t1:t2], pch = months, col = 1:4)


# setup plot 
years.f2009 <- 7
t1 <- 12*years.f2009 + 1
t2 <- t1 + 11

plot(time(detrended_xt_trn)[t1:t2], detrended_xt_trn[t1:t2], 
     type = "c", xlab = "", ylab = "", xaxt = 'n', lwd=2,
     main = "Detrended Residuals Throughout 2016") 
points(time(detrended_xt_trn)[t1:t2], detrended_xt_trn[t1:t2], pch = months, col = 1:4)

# setup plot 
years.f2009 <- 8
t1 <- 12*years.f2009 + 1
t2 <- t1 + 11

plot(time(detrended_xt_trn)[t1:t2], detrended_xt_trn[t1:t2], 
     type = "c", xlab = "", ylab = "", xaxt = 'n', lwd=2,
     main = "Detrended Residuals Throughout 2017") 
points(time(detrended_xt_trn)[t1:t2], detrended_xt_trn[t1:t2], pch = months, col = 1:4)

```

We now visualize the potential 40 months cycle.

```{r, fig.align='center', fig.height=6, fig.width=6}

# set up 3x1 block of plots 
par(mfrow=c(3, 1))

ts.plot(detrended_xt_trn[1:40], 
        type = "l", xlab = "Month", ylab="Detrended Residuals", 
        lwd = 2,
        main = "Detrended Residuals Throughout from 2009-2012") 

ts.plot(detrended_xt_trn[41:80], 
        type = "l", xlab = "Month", ylab="Detrended Residuals",
        lwd = 2, 
        main = "Detrended Residuals Throughout from 2012-2015") 

ts.plot(detrended_xt_trn[81:120], 
        type = "l", xlab = "Month", ylab="Detrended Residuals",
        lwd = 2, 
        main = "Detrended Residuals Throughout from 2015-2019") 
```


# Removing Cycles

Now we display the average annual cycle, and attempt to remove it.

```{r, warning=FALSE, message=FALSE, fig.height=7}

par(mfrow=c(2, 1))
ann.matrix <- matrix(detrended_xt_trn, byrow = TRUE, ncol = 12) 
ann.matrix[13, 12] <- NA # incomplete year in training cutoff

monthly_avg <- colMeans(ann.matrix, na.rm = TRUE) 
plot(1:12, monthly_avg, xlab = "Month", ylab = "Mean Value", 
     type = "o", 
     lwd=2, 
     main="Annual Cycle Average")
grid(lty="solid")
lines(1:12, monthly_avg, lwd = 2)

## Now remove this cycle and show the residuals. 
final_xt_trn <- detrended_xt_trn - monthly_avg


ts.plot(final_xt_trn, xlab = "Year", 
        ylab = "Remaining Residuals",
        main = "Detrended & Annual Cycle Removed", 
        type = "l", 
        lwd=2) 
grid(lty="solid")
lines(final_xt_trn, lwd = 2)


par(mfrow=c(1, 1))
# assess new spectrum
mvspec(final_xt_trn, 
      kernel("modified.daniell", c(2, 2)), # not a large set of data points
      taper=0.2,
      log="no", 
      lwd = 2)

```


Now we display the average 4 month cycle, and attempt to remove it.


```{r, warning=FALSE, message=FALSE, fig.height=7}

par(mfrow=c(2, 1))
four.month.matrix <- matrix(final_xt_trn, byrow = TRUE, ncol = 4) 
four.month.matrix[39, 4] <- NA # incomplete year in training cutoff

four.month.avg <- colMeans(four.month.matrix, na.rm = TRUE) 
plot(1:4, four.month.avg, 
     xlab = "# Month out of the Four", 
     ylab = "Mean Detrended Count",
     main = "Four-Month Cycle Average",
     type = "o", 
     lwd=2)
grid(lty="solid")
lines(1:4, four.month.avg, lwd = 2)

## Now remove this cycle and show the residuals. 
final_xt_trn <- final_xt_trn - four.month.avg

ts.plot(final_xt_trn, xlab = "Year", 
        ylab = "Remaining Residuals",
        main = "Detrended & Annual, Four-Month Cycles Removed", 
        type = "l", 
        lwd=2) 
grid(lty="solid")
lines(final_xt_trn, lwd = 2)


par(mfrow=c(1, 1))
# assess new spectrum
mvspec(final_xt_trn, 
      kernel("modified.daniell", c(2, 2)), # not a large set of data points
      taper=0.2,
      log="no",
      lwd = 2)
```


Now we display the average 40 month cycle, and attempt to remove it.


```{r, warning=FALSE, message=FALSE, fig.height=7}

par(mfrow=c(2, 1))
forty.month.matrix <- matrix(final_xt_trn, byrow = TRUE, ncol = 40) 
forty.month.matrix[4, c(36:40)] <- NA # incomplete year in training cutoff

forty.month.avg <- colMeans(forty.month.matrix, na.rm = TRUE) 
plot(1:40, forty.month.avg, 
     xlab = "# Month out of the 40", 
     ylab = "Mean Detrended Count", 
     main = "Forty-Month Cycle Average",
     lwd=2,
     type = "o")
grid(lty="solid")
lines(1:40, forty.month.avg, lwd = 2)

## Now remove this cycle and show the residuals. 
final_xt_trn <- final_xt_trn - forty.month.avg
  

ts.plot(final_xt_trn, 
        xlab = "Year", 
        ylab = "Remaining Residuals",
        main = "Detrended & All Cycles Removed", 
        lwd=2,
        type = "l") 
grid(lty="solid")
lines(final_xt_trn, lwd = 2)

par(mfrow=c(1, 1))
# assess new spectrum
mvspec(final_xt_trn, 
      kernel("modified.daniell", c(2, 2)), # not a large set of data points
      taper=0.2,
      log="no", 
      lwd=2)

```


# Estimate New ACF and PACF


```{r, warning=FALSE, message=FALSE}

# acf and pacf, before trend removed
acf2(final_xt_trn, max.lag = 20, 
     main="Series: Detrended & All Cycles Removed (Final) Residuals", 
     lwd = 2)

```


# Fitting the ARMA Model to Remaining Residuals

Note, the \texttt{arima()} function here uses conditional-sum-of-squares to find starting values, then applies maximum likelihood.


```{r, warning=FALSE, message=FALSE}

# loop through all 12x12 settings of ARMA(p, q) and calculate the AICs
aics <- matrix(0, nrow = 12, ncol = 12) 

for(i in 0:11) {
  for(j in 0:11) {
    
    if (i == 5 & j == 3){ 
      
      # issue with non-stationary for this combination of p and q (for CSS)
      t.arma <- arima(final_xt_trn, order = c(i, 0, j), method = "ML")
      aics[i+1,j+1] <- t.arma$aic
      
    } else {
      
      t.arma <- arima(final_xt_trn, order = c(i, 0, j))
      aics[i+1,j+1] <- t.arma$aic
    }

    
  }
} 

```


```{r, warning=FALSE, message=FALSE}

# load library for plot colors
library(RColorBrewer)

# setup colors
cols <- brewer.pal(12, "BrBG")


## AIC TUNING PLOT
par(mar=c(5,4,2,7))
plot(0:11, aics[, 1], 
     xlab = "AR Order (p)", 
     ylab = "AIC", 
     main = "Tuning ARMA(p, q) Grid-Search",
     type = "l", 
     lwd = 3,
     col = cols[1],
     ylim = c(1197, 1215))
grid(lty="solid")
for (r in 1:12) {
  
  lines(0:11, aics[, r], 
        lwd = 3,
        col = cols[r],
        type = "l")
  
}
legend("right", inset = c(-0.26,0), legend = 0:11, xpd = NA, 
       title = "MA Order (q)", col = cols, lty = 1, bty = "n", lwd=3)
points(3, min(aics), col="red", lwd=4, pch=19)


```


Now we re-fit the optimal ARMA(3, 8) model and display associated diagnostic plots and fitted coefficients (with full summary).


```{r}

# use sarima to get the diagnostics
sarima(final_xt_trn, p = 3, q = 8, d = 0)

# fit arima using regular arima model (same coefficients were found)
arma.mod <- arima(final_xt_trn, order=c(3, 0, 8))

# display coefficients and summary
arma.mod$coef

```


We now assess our model's fit to the training data. In addition, we display the associated AIC and training RMSE score.


```{r, warning=FALSE, message=FALSE}

# load library
library(forecast)

# overlay fitted to final residuals training data
ts.plot(final_xt_trn, 
        type = "l", 
        main = "Detrended + Cycles Removed Residuals with ARMA(3, 8) Fit",
        xlab = "Year", 
        ylab = "Detrended + Cycles Removed Residuals")
grid(lty="solid")
lines(final_xt_trn, lwd=2)
lines(list(x=time(final_xt_trn), y=fitted(arma.mod)), 
      lwd=2, 
      col="red")
legend("topleft", cex = 1, c("Actual", "Predicted"), 
       lty = 1, col = c("black", "red"), lwd=2)


# output AIC and training RMSE for residuals
cat("FOR RESIDUALS\n")
cat("AIC:", arma.mod$aic, " | RMSE:", sqrt(mean(arma.mod$residuals^2)), "\n\n")

# convert back to counts
trn.count.preds <- fitted(arma.mod) + monthly_avg + four.month.avg + 
                   forty.month.avg + lm.fit.d3$fitted.values

# overlay fitted to final residuals training data
ts.plot(xt_trn, 
        type = "l", 
        main = "Actual vs Predicted Counts for Training Data",
        xlab = "Year", 
        ylab = "Counts", 
        lwd=2)
grid(lty="solid")
lines(xt_trn, lwd=2)
lines(list(x=time(xt_trn), y=trn.count.preds), 
      lwd=2, 
      col="red")
legend("topleft", cex = 1, c("Actual", "Predicted"), 
       lty = 1, col = c("black", "red"), lwd=2)


# output training RMSE for counts
cat("FOR COUNTS\n")
cat("RMSE:", sqrt(mean((xt_trn-trn.count.preds)^2)), "\n\n")

```


# Spectrum of Residuals from ARMA(3, 8) Model


```{r, warning=FALSE, message=FALSE}

# assess spectrum residuals from ARMA
arma.res.spec <- mvspec(resid(arma.mod), 
                        log="no",
                        main="Series: ARMA(3, 8) Residuals | Smoothed Periodogram | L = 5", 
                        lwd=2)

# non-parametric
arma.res.spec.np <- mvspec(resid(arma.mod), 
                           kernel("modified.daniell", c(2, 2)),
                           log="no", 
                           main="Series: ARMA(3, 8) Residuals | Smoothed Periodogram | L = 5", 
                           lwd=2)


```


# Testing Performance with trained ARMA(3, 8)


```{r, warning=FALSE, message=FALSE, fig.height=5}
    

## TRAINING + TESTING COUNTS PLOTS 

# convert back to counts (+ cycles + trend)
tot.cnt.preds <- c(fitted(arma.mod), predict(arma.mod, 17)$pred) + 
                   monthly_avg + four.month.avg + forty.month.avg + 
                   predict(lm.fit.d3, data.frame(time=as.vector(time(xt))))

# overlay fitted to final residuals training data
ts.plot(xt, 
        type = "l", 
        main = "Actual vs Predicted Counts for Training + Testing Data",
        xlab = "Year", 
        ylab = "Counts")
grid(lty="solid")
lines(xt, lwd=2)
lines(list(x=time(xt), y=tot.cnt.preds), 
      lwd=2, 
      col="red")
rect(xleft=head(time(xt_tst),1), xright=tail(time(xt_tst),1),
     ybottom=par("usr")[3], ytop=par("usr")[4],
     density=NA, col = adjustcolor("blue", alpha = 0.15))
legend("topleft", cex = 1, c("Actual", "Predicted"), 
       lty = 1, col = c("black", "red"), lwd=2)


## ONLY TRAINING DATA PERFORMANCE

trnpreds <- tot.cnt.preds[1:(length(tot.cnt.preds)-18)]

# output training RMSE for counts
cat("Testing RMSE:", sqrt(mean((xt_trn - trnpreds)^2)), "\n\n")

## ONLY TESTING DATA PERFORMANCE

tstpreds <- tot.cnt.preds[(length(tot.cnt.preds)-17):length(tot.cnt.preds)]

# output testing RMSE for counts
cat("Testing RMSE:", sqrt(mean((xt_tst - tstpreds)^2)), "\n\n")

```


# Making Forecasts Beyond Available Data


```{r, warning=FALSE, fig.align='center', fig.height=10, fig.width=8}

# refit trend, and then remove
xt_df <- data.frame(time=as.vector(time(xt)), count=as.vector(xt))
t.fit <- lm(count ~ poly(time, 3), data=xt_df) # fit 
detrended_xt <- xt - t.fit$fitted.values # remove trend

## remove cycles again

ann.matrix <- matrix(detrended_xt, byrow = TRUE, ncol = 12) 
ann.matrix[15, c(6:12)] <- NA # incomplete year 2023
monthly_avg <- colMeans(ann.matrix, na.rm = TRUE) 

# remove annual cycle
final_xt <- detrended_xt - monthly_avg

four.month.matrix <- matrix(final_xt, byrow = TRUE, ncol = 4) 
four.month.matrix[44, c(2:4)] <- NA # incomplete year 2023
four.month.avg <- colMeans(four.month.matrix, na.rm = TRUE) 

# remove 4 month cycle
final_xt <- final_xt - four.month.avg

forty.month.matrix <- matrix(final_xt, byrow = TRUE, ncol = 40) 
forty.month.matrix[5, c(14:40)] <- NA # incomplete year 2023
forty.month.avg <- colMeans(forty.month.matrix, na.rm = TRUE) 

# remove 40 month cycle
final_xt <- final_xt - forty.month.avg


# fit arima using regular arima model (same coefficients were found)
full.arma.mod <- arima(final_xt, order=c(3, 0, 8))

## FORECASTING PLOT

# make forecast
fore <- predict(full.arma.mod, 36)

fore.preds <- ts(start = c(2023, 1), end = c(2026, 5), frequency = 12)
fore.preds[-c(1:5)] <- fore$pred

# generate fitted values
tot.fitted <- fitted(full.arma.mod) + 
                     monthly_avg + four.month.avg + forty.month.avg + 
                     fitted(t.fit)

# convert to counts (have to align starting at Jan to recycle with the 
# averages correctly)
fore.preds <- fore.preds +
              monthly_avg + four.month.avg + forty.month.avg + 
              predict(t.fit, data.frame(time=as.vector(time(fore.preds))))

# remove NAs
fore.preds <- window(fore.preds, start=c(2023, 6))

# for CI 
U <- fore.preds + fore$se
L <- fore.preds - fore$se
xx <- c(time(U), rev(time(U)))
yy <- c(L, rev(U))

# overlay fitted to final residuals training data
par(mfrow=c(2, 1))
ts.plot(xt, 
        type = "l", 
        main = "Actual vs Predicted Counts (w/ Forecasting)",
        xlab = "Year", 
        ylab = "Counts", 
        xlim=c(2009, 2026.3))
grid(lty="solid")
lines(xt, lwd=2)
lines(list(x=time(xt), y=tot.fitted), 
      lwd=2, 
      col="red")
polygon(xx, yy, border = 8, col = gray(.6, alpha = .2))
lines(list(x=time(fore.preds), y=fore.preds), lwd=2, col="blue")
legend("topleft", cex = 0.8, c("Actual", "Fitted", "Forecast"), 
       lty = 1, col = c("black", "red", "blue"), lwd=2)

ts.plot(xt, 
        type = "l",
        xlab = "Year", 
        ylab = "Counts", 
        main = "Forecast Enhanced",
        lwd = 2,
        xlim=c(2021.5, 2026.1), 
        ylim=c(90, 220))
grid(lty="solid")
lines(xt, lwd = 2)
lines(list(x=time(xt), y=tot.fitted), 
      lwd=3, 
      col="red")
polygon(xx, yy, border = 8, col = gray(.6, alpha = .2))
lines(list(x=time(fore.preds), y=fore.preds), lwd=2, col="blue")
legend("topright", cex = 0.8, c("Actual", "Fitted", "Forecast"), 
       lty = 1, col = c("black", "red", "blue"), lwd=3)



```

